Ordinal Classification Using Partially Ordered Feature Sets

### 2025-02-28 paper/main.tex
Add files via upload

\journal{Decision Support Systems}

@@

\begin{abstract}
This paper presents a method to solve an ordinal classification problem where the objects are described with a set of partially ordered features.
To construct an ordinal classification model we aggregate partial orders of features by weighting the incidence matrices of partial order graphs. To describe set of possible solutions we use partial order cones corresponding to the incidence matrices and find a solution of the classification problem as the projection of a vector of response variables to a superposition of the cones. We propose a method of optimum parameter estimation for the ordinal classification model. The method is applied to the Football Player List categorization problem.
\end{abstract}

@@

The proposed method is compared with oblique decision tree~\cite{Wilhelm2021} and isotonic regression~\cite{SALMON2017} on synthetic and real data. As a real data we consider a problem of the Football Player List categorization, also described in. The goal is to rank FIFA players by their world reputation ranks basing on their characteristics. The following categories are considered: least popular, rarely heard, normally heard, popular, extremely popular. This categorization is monotonic. The partially ordered features (e.g., mobility, strength) are given by the experts. The proposed method shows better results in comparison with the alternative approaches.

\begin{table}[!htbp]
\label{tab:intro_comparative}
\caption{Comparative analysis of basic solution to my problem.}
\begin{tabular}{p{5cm}|p{5cm}|p{5cm}}
	Solution & Strengths & Weakness \\
	\hline
	Partial Orders & 
	no contradictions in the partial order &
	vulnerable to noise \\
    \hline
	Isotonic regression & 
	working fast &
	inaccurately predictions \\
    \hline
	Oblique decision trees & 
	learns non-linearity dependence &
	vulnerable to retraining \\
    \hline
	Stochastic Partial Orders & 
	no vulnerability to noise &
	demand higher dimension of features \\
\end{tabular} 
\end{table}

\section{Ordinal classification with monotonicity constraints}
\paragraph{Partially ordered feature set}
Let~$\mathfrak{D}$ be a sample consisting of the pairs
\[
\mathfrak{D}=\{(\bx_i,y_i)\}_{i=1}^m,
\label{eq:Sample}
\]
where~$\bx_i=[x_{i1},...,x_{ij},...,x_{in}]\T$~is an object to be classified, $y_i$~is a class label. An object~$\bx_i$ is an $n$-dimensional vector, each component of which belongs to the corresponding partially ordered set~$X_j$. In other words, an element~$j$ of the vector~$\bx_i$ (which is also reffered to as value of the feature~$j$ for the object~$\bx_i$) belongs to the set~$X_j$ with the given partial order relation~$\succeq$ with the following properties:
\begin{itemize}
\item reflexivity, $\forall a\in X$ $(a\succeq a)$,
\item antisymmetry, $\forall a,b\in X$, $(a\succeq b)\wedge(b\succeq a)\Rightarrow (a= c)$,
\item transitivity, $\forall a,b,c\in X$ $(a\succeq b)\wedge(b\succeq c)\Rightarrow (a\succeq c)$.
\end{itemize}
Thereby the object set~$X$ is a Cartesian product of the partially ordered sets~$X_1,...,X_n$,
\[
X=X_1\times X_2\times \dots \times X_n,
\]
the sets~$X_1,...,X_n$ are the feature values sets.

@@

For each set~$X_j$ define a cone~$\mathcal{X}_j$ in a space~$\mathbb{R}^m_+$:
\begin{equation}
\mathcal{X}_j=\{\bchi_j\in\mathbb{R}^m_+|\quad x_{ij}\succeq x_{kj}\rightarrow \chi_{ij}\geq \chi_{kj}\quad \forall i,k=1,...,m\},
\label{ConeDef}
\end{equation}
where~$x_{ij}$~and~$x_{kj}$~are values of the feature~$j$ on the sample elements~$i$~and~$k$, respectively.
For the cone~$\mathcal{X}$ (index~$j$ is omitted for the reasons of convenience) the following theorem holds.
\begin{Theorem}
A vector~$\bchi$ belonging to the cone~$\mathcal{X}$ can be represented as the nonnegative combination of cone generators,
$$\bchi=\sum\limits_{k=1}^{m}\lambda_{k}\bzeta_{k}, \quad\lambda_{k}\geq 0,$$
where~$\bzeta_k$~is a generatrix of the cone~$\mathcal{X}$,
\[
\bzeta_{k}(i)=
\begin{cases}
1,\quad \text{if}\quad x_i\succeq x_k,\\
0,\quad\text{if}\quad x_i\nsucceq x_k,
\end{cases}
\]
and this decomposition is unique.

@@

We examine decision trees that test a linear combination of the attributes
at each internal node. More precisely, let an example take the form $X = x1; x2; \dots xd$; $C_j$
where $C_j$ is a class label and the $x_i$ are real-valued attributes. The test at each node will
then have the form:

@@

To estimate parameters of the model~(\ref{ordClassMdl}) we propose a two-stage algorithm, like we have said before. In the first stage we estimate the partial order matrix~$\bPsi$, where the elements of the matrix~$\bPsi$ equal to the values of function~$\Psi$ on the objects of the sample~$\mathfrak{D}$,
\[
\bPsi(i,k) = \Psi(\bx_i,\bx_k).
\]
In the second stage estimate parameters~$\lambda_k$ of a linear combination~$\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)$ from~(\ref{utilityDecompDual}).
\paragraph{Partial order matrix estimation}
Estimate matrix~$\bPsi$ of the pairwise objects dominance using initial approximation as the matrix P. By matrix P we consider
$$P_0 = \begin{pmatrix} P(x_1 \succeq x_1 | \mu_1 \succeq \mu_1) & P(x_1 \succeq x_2 | \mu_1 \succeq \mu_2) &                       &                       &    \\
  & \ddots  & P(x_i \succeq x_j | \mu_i \succeq \mu_j)                    &                       &    \\
  &    & \ddots & \ddots &    \\
  &    &                       &                     & P(x_n \succeq x_n | \mu_n \succeq \mu_n)\end{pmatrix}
 \in \mathbb{R}^{n \times n}\enspace,
$$
This matrix can be easily calculated by using sampling methods.
Let every element of a matrix~$\bPsi$ be a linear combination of elements of the matrices~$\bZ_1,...,\bZ_n$, according to~(\ref{problLinearMon}):
\[
\bPsi(i,k)=\sum\limits_{j=1}^n w_j\bZ_j(i,k).
\]
Here we have to assume the probability of existing the noise within sample data. So it is necessary to use learning without teacher - that is how will be used matrix P.
The optimal parameters~$\hat{\bw}$ minimize a loss function
\[
\hat{\bw} = \arg\min\limits_{\bw}\sum\limits_{i=1}^m\sum\limits_{k=1}^m\left(P_0(i,k) - \sum\limits_{j=1}^n w_j\bZ_j(i,k)\right)^2,
\]
here~$\bP$~is considered to be partial order matrix corresponding to the class labels vector~$\by$.
Since the target variable~$P_0(i,k)$ is $\in[0,1]$, we propose to estimate an probability distribution vector~$\hat{\bw}$ using one of the standard optimization methods. In this paper we use the Mean Least Squares method.

@@

\begin{table}[h]
\centering
\caption{Algorithm comparison on the Football Player dataset. The loss function is Hamming distance between class labels.}
\begin{tabular}{|c|c|c|}
  \hline
  Algorithm & Learn error & Test error \\
  \hline
  Partial Orders & 1.14 & 1.69 \\
  \hline
  Isotonic Regression & 0.98 & 1.28 \\
  \hline
  Oblique Decision Trees & 0.47 & 1.06 \\
  \hline
  Stochastic Partial Orders & 1.48 & 1.32 \\
  \hline
\end{tabular}
\label{tab:AlgComparison}
\end{table}

\section{Conclusion}
The proposed method of the ordinal classification problem uses partially ordered sets of expert estimations of features. We considered an alternative problem statement for the ordinal classification with monotonic constraints. Each partially ordered feature set corresponds to the partial order cone and to the matrix of the partial order graph. We find the solution of an ordinal classification problem as the projection to the cones superposition. The proposed method is compared with alternative approaches on the Football Player List dataset. The quality of classification was improved using the ideas considered in the present paper.

>>>>

\journal{Expert Systems}

@@

\begin{abstract}
This paper presents a method to solve an ordinal classification problem where the objects are described with a set of partially ordered features.
To construct an ordinal classification model we aggregate partial orders of features by weighting the incidence matrices of partial order graphs. To describe set of possible solutions we use partial order cones corresponding to the incidence matrices and find a solution of the classification problem as the projection of a vector of response variables to a superposition of the cones. We propose a method of optimum parameter estimation for the ordinal classification model. The method is applied to the Football Player List, Student Marks List and Red List categorization problems.
\end{abstract}

@@

The proposed method is compared with oblique decision tree~\cite{Wilhelm2021} and isotonic regression~\cite{SALMON2017} on synthetic and real data. As a basic example of real data we consider a problem of the Football Player List categorization, also described in. The goal is to rank FIFA players by their world reputation ranks basing on their characteristics. The following categories are considered: least popular, rarely heard, normally heard, popular, extremely popular. This categorization is monotonic. The partially ordered features (e.g., mobility, strength) are given by the experts. The proposed method shows better results in comparison with the alternative approaches.

\begin{table}[!htbp]
\label{tab:intro_comparative}
\caption{Comparative analysis of basic solution to my problem.}
\begin{tabular}{|p{5cm}|p{5cm}p{5cm}|} \hline
	Algorithm & Strengths & Weakness \\
	\hline
	Partial Orders & 
	no contradictions in the partial order &
	vulnerable to noise \\
    \hline
	Isotonic regression~\cite{SALMON2017} & 
	working fast &
	inaccurately predictions \\
    \hline
	Oblique decision trees~\cite{Wilhelm2021} & 
	learns non-linearity dependence &
	vulnerable to retraining \\
    \hline
	Stochastic Partial Orders & 
	no vulnerability to noise &
	demand higher dimension of features \\
    \hline
\end{tabular} 
\end{table}

Stochastic Partial Orders method still in the development cause the limits of the data. Such task, like solving ranking-categorization problem is specific and it may be difficult to find appropriate datasets for that. We assuming that by the time, with appearing of data with higher dimension of features, this new method will perform much better results.

\section{Ordinal classification with monotonicity constraints}
\paragraph{Partially ordered feature set}
Let~$\mathfrak{D}$ be a sample consisting of the pairs
\[
\mathfrak{D}=\{(\bx_i,y_i)\}_{i=1}^m,
\label{eq:Sample}
\]
where~$\bx_i=[x_{i1},...,x_{ij},...,x_{in}]\T$~is an object to be classified, $y_i$~is a class label. An object~$\bx_i$ is an $n$-dimensional vector, each component of which belongs to the corresponding partially ordered set~$X_j$. In other words, an element~$j$ of the vector~$\bx_i$ (which is also reffered to as value of the feature~$j$ for the object~$\bx_i$) belongs to the set~$X_j$ with the given partial order relation~$\succeq$ with the following properties:
\begin{itemize}
\item reflexivity, $\forall a\in X$ $(a\succeq a)$,
\item antisymmetry, $\forall a,b\in X$, $(a\succeq b)\wedge(b\succeq a)\Rightarrow (a= b)$,
\item transitivity, $\forall a,b,c\in X$ $(a\succeq b)\wedge(b\succeq c)\Rightarrow (a\succeq c)$.
\end{itemize}
Thereby the object set~$X$ is a Cartesian product of the partially ordered sets~$X_1,...,X_n$,
\[
X=X_1\times X_2\times \dots \times X_n,
\]
the sets~$X_1,...,X_n$ are the feature values sets.

@@

For each set~$X_j$ define a cone~$\mathcal{X}_j$ in a space~$\mathbb{R}^m_+$:
\begin{equation}
\mathcal{X}_j=\{\bchi_j\in\mathbb{R}^m_+|\quad x_{ij}\succeq x_{kj}\rightarrow \chi_{ij}\geq \chi_{kj}\quad \forall i,k=1,...,m\},
\label{ConeDef}
\end{equation}
where~$x_{ij}$~and~$x_{kj}$~are values of the feature~$j$ on the sample elements~$i$~and~$k$, respectively.
For the cone~$\mathcal{X}$ (index~$j$ is omitted for the reasons of convenience) the following theorem holds.
\begin{Theorem}
A vector~$\bchi$ belonging to the cone~$\mathcal{X}$ can be represented as the nonnegative combination of cone generators,
$$\bchi=\sum\limits_{k=1}^{m}\lambda_{k}\bzeta_{k}, \quad\lambda_{k}\geq 0,$$
where~$\bzeta_k$~is a generatrix of the cone~$\mathcal{X}$,
\[
\bzeta_{k}(i)=
\begin{cases}
1,\quad \text{if}\quad x_i\succeq x_k,\\
0,\quad\text{if}\quad x_i\nsucceq x_k,
\end{cases}
\]
and this decomposition is unique.
\end{Theorem}

@@

We examine decision trees that test a linear combination of the attributes
at each internal node. More precisely, let an example take the form $X = x_1; x_2; \dots x_d$; $C_j$
where $C_j$ is a class label and the $x_i$ are real-valued attributes. The test at each node will
then have the form:

@@

To estimate parameters of the model~(\ref{ordClassMdl}) we propose a two-stage algorithm, like we have said before. In the first stage we estimate the partial order matrix~$\bPsi$, where the elements of the matrix~$\bPsi$ equal to the values of function~$\Psi$ on the objects of the sample~$\mathfrak{D}$,
\[
\bPsi(i,k) = \Psi(\bx_i,\bx_k).
\]
In the second stage estimate parameters~$\lambda_k$ of a linear combination~$\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)$ from~(\ref{utilityDecompDual}).
\paragraph{Partial order matrix estimation}
Estimate matrix~$\bPsi$ of the pairwise objects dominance using initial approximation as the matrix P. By matrix P we consider
$$P_0 = \begin{pmatrix} P(x_1 \succeq x_1 | \mu_1 \succeq \mu_1) & P(x_1 \succeq x_2 | \mu_1 \succeq \mu_2) &                   \\
 \ddots   & P(x_i \succeq x_j | \mu_i \succeq \mu_j) & \cdots                                  \\
  &    & \ddots     \\
  &    &                    P(x_n \succeq x_n | \mu_n \succeq \mu_n)\end{pmatrix}
 \in \mathbb{R}^{n \times n}\enspace,
$$
This matrix can be easily calculated by using sampling methods.
Let every element of a matrix~$\bPsi$ be a linear combination of elements of the matrices~$\bZ_1,...,\bZ_n$, according to~(\ref{problLinearMon}):
\[
\bPsi(i,k)=\sum\limits_{j=1}^n w_j\bZ_j(i,k).
\]
Here we have to assume the probability of existing the noise within sample data. So it is necessary to use learning without teacher - that is how will be used matrix P.
The optimal parameters~$\hat{\bw}$ minimize a loss function
\[
\hat{\bw} = \arg\min\limits_{\bw}\sum\limits_{i=1}^m\sum\limits_{k=1}^m\left(P_0(i,k) - \sum\limits_{j=1}^n w_j\bZ_j(i,k)\right)^2,
\]
here~$\bP$~is considered to be partial order matrix corresponding to the class labels vector~$\by$.
Since the target variable~$P_0(i,k)$ is $\in[0,1]$, we propose to estimate an probability distribution vector~$\hat{\bw}$ using one of the standard optimization methods. In this paper we use the Mean Least Squares method.

@@

\begin{table}[h]
\centering
\caption{Algorithm comparison on the Football Player dataset. The loss function is Hamming distance between class labels.}
\begin{tabular}{|c|cc|}
  \hline
  Algorithm & Learn error & Test error \\
  \hline
  Partial Orders & 1.14 $\pm$ 0.05 & 1.69 $\pm$ 0.2 \\
  \hline
  Isotonic Regression & 0.98 $\pm$ 0.2 & 1.28 $\pm$ 0.4 \\
  \hline
  Oblique Decision Trees & 0.47 $\pm$ 0.5 & 1.06 $\pm$ 0.71 \\
  \hline
  Stochastic Partial Orders & 1.48 $\pm$ 0.14 & 1.32 $\pm$ 0.05 \\
  \hline
\end{tabular}
\label{tab:AlgComparison}
\end{table}

\section{Conclusion}
The proposed method of the ordinal classification problem uses partially ordered sets of expert estimations of features. We considered an alternative problem statement for the ordinal classification with monotonic constraints. Each partially ordered feature set corresponds to the partial order cone and to the matrix of the partial order graph. We find the solution of an ordinal classification problem as the projection to the cones superposition. The proposed method is compared with alternative approaches on the varios examples of datasets: such as Football Player List, Student Marks List and Red List. The quality of classification was improved using the ideas considered in the present paper.

\section{Availability of data and materials}

All datasets and software used for supporting the conclusions of this article are available form the public data repository at the website of https://github.com/papayiv/Papay-BS-Thesis/tree/main/code.

\section{Authors' contributions}

VVS designed, coordinated this research and drafted the manuscript. IDP carried out experiment and data analysis, conceived of the study and participated in research coordination. The authors read and approved the final manuscript.
